---
title: "Titanic data exploration markdown"
author: "Olivia Sheng"
date: "May 12, 2020"
output: 
  html_document:
    number_sections: yes
    theme: readable
    highlight: breezedark
    toc: yes
    fig_width: 25
    fig_height: 15
editor_options: 
  chunk_output_type: console
---
# Set up, data import, inspection and transformation
## Set up, data import and inspection

```{r Set up, data import and inspection}
# The sinking of the RMS Titanic is one of the most infamous shipwrecks in history.
# On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with 
# an iceberg, killing 1502 out of 2224 passengers and crew.
# This sensational tragedy shocked the international community and led to better 
# safety regulations for ships.One of the reasons that the shipwreck led to such 
# loss of life was that there were not enough lifeboats for the passengers and crew. 
# Although there was some element of luck involved in surviving the sinking, 
# some groups of people such as women, children, and the upper-class 
# were more likely to survive than others.

# VARIABLE DESCRIPTIONS:

# PassengerID     Unique passenger identifier
# Survived        Survival (0 = No; 1 = Yes)
# Pclass          Passenger Class(1 = 1st; 2 = 2nd; 3 = 3rd) (Pclass is a proxy for socio-economic status (SES)
#                     1st ~ Upper; 2nd ~ Middle; 3rd ~ Lower)
# Name            Name
# Sex             Sex
# Age             Age (Age is in Years; Fractional if Age less than One (1) If the Age is Estimated, it is in the form xx.5)
# Sibsp           Number of Siblings/Spouses Aboard
# Parch           Number of Parents/Children Aboard
# Ticket          Ticket Number
# Fare            Passenger Fare
# Cabin           Cabin
# Embarked        Port of Embarkation (C = Cherbourg; Q = Queenstown; S = Southampton)


### ---------------------------------------------------------------------------

### Import a csv file
library(tictoc) 
tic()

cloud_wd <- getwd()
setwd(cloud_wd)

titanic <- read.csv(file = "titanic.train.csv", stringsAsFactors = FALSE)

### Examine the overall data frame

# str() shows the number of observations, and the number, names, types and some values of columns

str(titanic)

## You can retrieve and save the number of rows and number of coloumns of a data frame

nrow(titanic)
row <- nrow(titanic)
row

col <- ncol(titanic)
col

## Show the head and tail rows of a data frame

head(titanic)

titanic[1:6,]

head(titanic, n=1)

titanic[1,]

head(titanic, n=10)

tail(titanic, n = 10)

?head

# summary() shows the mean and the five-number statistics indicating the spread of each column's values

summary(titanic)
```
## Data transformation
Remove unique identifiers from further analysis as they are not interesting without additional feature extractions

```{r Data transformation 1}
titanic <- titanic[c(-1,-4,-9)]
```
Change Survived and other nominal variables to factors
```{r Data transformation 2}
titanic$Survived <- factor(titanic$Survived)

titanic$Sex <- factor(titanic$Sex)

titanic$Pclass <- factor(titanic$Pclass)

titanic$Cabin <- factor(titanic$Cabin)

titanic$Embarked <- factor(titanic$Embarked)

str(titanic)

summary(titanic)
```
Remove observations with missing Age values. 

This missing data handling approach has the obvious disadvantages of 
the applicability of the model to data with missing age.

To keep observations with missing Age values require careful imputation of Age missingness. 

The various missing data imputation methods are beyond the knowledge required for this tutorial.

```{r Data transformation 3}
# Missing values in Age can cause problems. Test the following commands when you have time.
# var(titanic$Age)
# mean(titanic$Age)
# var(titanic[-which(is.na(titanic$Age)), ]$Age)
# sd(titanic[-which(is.na(titanic$Age)), ]$Age)

# Use which() command to identify and remove observations in which 
# the Age value is missing or is.na(titanic Age) = TRUE.

titanic <- titanic[-which(is.na(titanic$Age)), ]

?which

summary(titanic)
```

On the other hand, keeping missing factor levels might be able to lead to meaningful models
Empty level names of the Cabin and Embarked factors will cause problems in some analysis.
Other missing factor value imputation remains a good option beyond the scope of this tutorial.


```{r Data transformation 4}
# levels() and sort() with factors
str(titanic$Cabin)

titanic$Cabin

# str() truncate the display of factor levels. levels() shows all of the factor levels of a factor variable

levels(titanic$Cabin)

?levels

# It is useful to sort the levels based on how many observations contain a factor level.
# Remember table() that counts how many observations contain a factor level.
# Use sort to sort factor levels by the count of observations containing a level
# You can use sort inside barplot()\

table(titanic$Cabin)

sort(table(titanic$Cabin), decreasing = TRUE)

?sort

barplot(sort(table(titanic$Cabin), decreasing = TRUE))

barplot(sort(table(titanic$Cabin), decreasing = FALSE))

# Fixing empty character level names for Cabin and Embarked

levels(titanic$Cabin)[1]

levels(titanic$Cabin)[1] <- "missing"

levels(titanic$Embarked)

barplot(sort(table(titanic$Embarked), decreasing = FALSE))

levels(titanic$Embarked)[1]

levels(titanic$Embarked)[1] <- "missing"

summary(titanic)
```
# Understanding numeric variables

Show summary of one or more columns
```{r Understanding numeric variables}
summary(titanic$Pclass)

summary(titanic[c("Sex", "Age")])

## obtain the mean, median, max, min and range of a numeric variable

mean(titanic$Age)
median(titanic$Age)
range(titanic$Age)

# calculate the difference of the range

max.Age <- max(titanic$Age)
min.Age <- min(titanic$Age)
min.Age
max.Age

range.diff.Age <- max.Age - min.Age
range.diff.Age


# min-max normalization of first observation's Age to a value between zero zna 1

titanic$Age[1]

(titanic$Age[1] - min.Age)/range.diff.Age

# diff() calculates differences between an attribute's values in the referenced record and that in the record following it
# diff(titanic$Age) 
# ?diff
# titanic$Age

# use quantile to calculate the five-number summary for Age

quantile(titanic$Age)

# the 1st percentile and the 99th percentile

quantile(titanic$Age, probs = c(0.01, 0.99))

# quintiles and deciles 

quantile(titanic$Age, seq(from = 0, to = 1, by = 0.20))

quantile(titanic$Age, seq(from = 0, to = 1, by = 0.10))

# boxplot of a numeric variable

boxplot(titanic$Age, main="Boxplot of Age in the titanic data set",
        ylab="Age")

# practice quantile for another variable - SibSp

quantile(titanic$SibSp)

# the 1st percentile and the 99th percentile

quantile(titanic$SibSp, probs = c(0.01, 0.99))

# quintiles and deciles 

quantile(titanic$SibSp, seq(from = 0, to = 1, by = 0.20))

quantile(titanic$SibSp, seq(from = 0, to = 1, by = 0.10))

# boxplot of a numeric variable

boxplot(titanic$SibSp, main="Boxplot of Sibsp in the titanic data set",
        ylab="Sibsp")
?IQR

### IQR = 3rdQaurtile - 1stQuartile

### maxline = 3rdQuartile + 1.5*IQR

### minline = 1stQuartile - 1.5*IQR

### one definition of outliers : value > maxline and value < minline are drawn as circles and dots 

# scatter plot of a numeric variable. It becomes too busy for a large number of observations

plot(titanic$SibSp)

# histograms of a numeric variable

hist(titanic$SibSp, main = "Histogram of Sibsp in the titanic data set",
     xlab = "Sibsp")

### For a right skewed distribution, the mean is typically greater than the median

hist(titanic$Age, main = "Histogram of Age in the titanic data set",
     xlab = "Age")

hist(titanic$Fare, main = "Histogram of Fare in the titanic data set",
     xlab = "Fare")

hist(titanic$Parch, main = "Histogram of Parch  in the titanic data set",
     xlab = "Parch")

### Search for and understand the meaning of skewed, bimodal and multimodal continuous distributions

## variance and standard deviation of a numeric value

var(titanic$Age)

sd(titanic$Age)

var(titanic$SibSp)

sd(titanic$SibSp)
```
# Exploring factor variables
```{r Explorating factor variables}
# A factor's distinct values

is.factor(titanic$Survived)

summary(titanic$Survived)

nlevels(titanic$Survived)

str(titanic$Survived)

is.factor(as.character(titanic$Survived))

summary(as.character(titanic$Survived))

# How many of observations have a factor value?

?table

plot(titanic$Survived)

plot(titanic$Survived, main = "Plot of Survived in the titanic data set",
     xlab = "Survived")

Survived.table <- table(titanic$Survived)

Survived.table

str(Survived.table)

barplot(Survived.table, main = "Plot of Survived in the titanic data set",
        xlab = "Survived")
# remember the difference between the input data structures to plot() and barplot()

# compute table proportions

? prop.table

# Run prop.table(titanic$Survived) to see the error in this command

prop.table(Survived.table)

Survived.prop <- prop.table(table(titanic$Survived))

Survived.prop 

## round decimals 

round(Survived.prop, digits = 2)

options(digits = 2)

prop.table(Survived.prop)

###Remeber to get help from Help Pane using "?"

?round
?options
```
# Understand relatioinships of multiple variables
```{r Understand variable relationships }
# cor,  boxplot, 2D scatter plot - plot, 3D scatter plot

# scatter plot: two numeric variables

plot(titanic[,5:6])
plot(titanic$SibSp, titanic$Parch)

### Generate correlation coefficients of two numeric variables in a 2x2 matrix
# cor(X,Y) lies between -1 and 1. zero means no correlation. 1 or -1 indicates full correlation
# positive value means positive correlation and negative values mean negative relationships
# Examine the components in the formulation for correlation coefficients
# cor(X,Y) = cov(X,Y)/(sd(X)*sd(Y)) 
# cov(X,Y) = E[X-E(X)]*E[Y-E(Y)]

cov(titanic[,c(5,6)])
var(titanic[,c(5,6)])
var(titanic[,5])
sd(titanic[,5])
var(titanic[5])
sd(titanic[,6])
cor(titanic[c("SibSp", "Parch")])
cor(titanic[5:6])
cor(titanic[,5:6])
?cor

## Generate the correlation matrix of all numeric variables

cor(titanic[4:7])

# Generate 2D scatter plots and correlation coefficients

# Load the psych package. Install it first when necessary.

pairs(titanic[4:7])

library(psych)

pairs.panels(titanic[-1])

pairs.panels(titanic)
```
# Examine relationships between numeric variables and factors

##boxplot groups values of a numeric variable based on the values of a factor
```{r boxplot numeric values by factors}
boxplot(Age~Survived, data = titanic)
boxplot(Parch~Survived, data = titanic)
boxplot(SibSp~Survived, data = titanic)
boxplot(Fare~Survived, data = titanic)
```
## The aggregate function
```{r Aggregation function}
# We can use the aggregate command to aggregate a numeric feature by a categorical one.

# The aggregate function has three parameters

# 1. The numeric value, e.g. sales, to be aggregated to find out, e.g., total of sales,
#   average of sales, number of sales (i.e. orders).

# 2. The set of categories, product_category and sales_region, on which you wish
#   to aggregate

# 3.The aggregation function (e.g., sum, mean, length) that you wish to use

aggregate(SibSp~Survived, summary, data = titanic)

aggregate(Parch~Survived, summary, data = titanic)

```
## scatter plot of numeric values and factor values
```{r Scatter plat of numeric and factor variables}
plot(titanic$Survived)
plot(titanic$Age)
plot(titanic$Age,titanic$Survived)

plot(titanic$SibSp,titanic$Parch, col=titanic$Survived, pch = as.numeric((titanic$Survived)))

with(titanic, plot(titanic$SibSp,titanic$Parch, col=Survived, pch = as.numeric(Survived)))

with(titanic, plot(titanic$SibSp,titanic$Parch, col=Survived, pch = as.numeric(Survived), 
                   main = "2d scatter plot of titanic data", sub = "SibSp vs Parch"))


palette()

legend('topright', legend = levels(titanic$Survived),  col = 1:2, cex = 0.8, pch = 1:2)

```
## Generate 3D scatterplot
```{r Generate 3D scatterplot}
library(scatterplot3d)

scatterplot3d(titanic$Sex,titanic$Age,titanic$SibSp, pch = as.numeric(titanic$Survived), main = "3D scatter plot of titanic data")

legend('topright', legend = levels(titanic$Survived),  cex = 0.8, pch = 1:2)

scatterplot3d(titanic$Fare,titanic$Pclass,titanic$Sex, pch = as.numeric(titanic$Survived), main = "3D scatter plot of titanic data")

legend('topright', legend = levels(titanic$Survived),  cex = 0.8, pch = 1:2)

titanicf <- subset(titanic, Sex == "female")
summary(titanicf)
titanicm <- subset(titanic, Sex == "male")
summary(titanicm)

plot(titanicm$Age,titanicm$SibSp, pch = as.numeric(titanicm$Survived), main = "2D scatter plot of males' SibSp and Age")

legend('topright', legend = levels(titanicf$Survived),  cex = 0.8, pch = 1:2)

plot(titanicf$Fare,titanicf$Pclass, pch = as.numeric(titanicf$Survived), main = "2D scatter plot of titanic females' Pclass vs Fare")

legend('topright', legend = levels(titanicf$Survived),  cex = 0.8, pch = 1:2)
```

## Parallel plots
Another parallel plot and parallel coordinate plot to show relationships between numeric variables and factors

load the required libraries - lattice and MASS

Install these package first or use require(lattice) and require(MASS) to install and load them

```{r Generate parallel plots}
library(lattice)


parallelplot(~titanic[-1] | Survived, data = titanic, var.label = TRUE)

parallelplot(~titanic[4:7] | Survived, data = titanic, var.label = TRUE)

parallelplot(~titanic[4:5] | Survived, data = titanic, var.label = TRUE)

parallelplot(~titanic[c(4,6)] | Survived, data = titanic, var.label = TRUE)

parallelplot(~titanic[c(6,7)] | Survived, data = titanic, var.label = TRUE)

parallelplot(~titanic[c(5,6)] | Survived, data = titanic, var.label = TRUE)


palette()

# legend() that works for other plots cannot display legend in the parallelplot() above
```
## Generate a parallelcoordinate plot
```{r Generate a parallel coordinate plot}
library(MASS) 

parcoord(titanic[4:7], col = titanic$Survived, var.label = TRUE)

parcoord(titanic[4:5], col = titanic$Survived, var.label = TRUE)

parcoord(titanic[c(4,6)], col = titanic$Survived, var.label = TRUE)

parcoord(titanic[c(4,7)], col = titanic$Survived, var.label = TRUE)

parcoord(titanic[c(5,6)], col = titanic$Survived, var.label = TRUE)

palette()

legend('topright', legend = levels(titanic$Survived),  col = 1:2, cex = 0.8, pch = 1:2)

##### end of Titanic Data Exploration Tutorial
toc()
```
